The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-06-20 paper.qmd
finished general restructure for signal supression perspective

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
- Update background to reflect new structure of paper
-->

@@

There are many stages in our analysis that might benefit from the power of data visualisation, however this does not mean it is appropriately utilized or done with with effectiveness. Unlike an infographic, a statistical graphic is one of intent, and all enhancements made on a statistical graphic should be in pursuit of that goal (@Vanderplas2020). Visualization is an important step in exploratory data analysis and communication. At the exploration stage, it is utilized to **learn** what is important about a data set while the communication stage is utilized to **tell** others about what we have learned. Each of these visualisation goals utilize an important attribute of visualisation. In both these cases, visualisations are powerful tools that can help us understand **how** our data or results might diverge from pre-conceived notions or what we expect.

@@

The review provided here attempts to address these issues. We start by discussing why uncertainty visualisation has become it's own sub-field, distinct from "normal" information visualisation. In this discussion we establish the purpose of uncertainty visualisation is to perform "signal-suppression", that is, hide statistically insignificant signals while promoting statistically rigorous signals. **EDIT FROM HERE** We then move highlight the three key issues facing the uncertainty visualisation literature, specifically the lack of a definition of uncertainty, the confusion as to what *is* an "uncertainty visualisation", and the difficulties with measuring "uncertainty". When disucssing each of these issues, we will discuss how this concept is understood in existing literature, why this understanding may be flawed, and the issues that arrise due to these flaws. By considering the *purpose* of visualizing uncertainty we avoid providing a simple list of the current methods, and instead consider if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology. The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarizing different viewpoints the current literature, and provide a guide for potential research topics. **TO HERE**

# The purpose of uncertainty visualisation
It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*. A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic. Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, however this field *does* implicitly treat them different. Why?

Uncertainty visualisation papers usually provide one of two primary justifications for their existence. The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty. Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019]. The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate. Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].

@@

# Signal-Supression
The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.

@@

## Uncertainty does not exist without signal
It is obvious that we cannot design a visualisation for signal suppression without a clear signal. This may be considered a limitation by some readers, however it is less of a restriction when we recognize that *uncertainty* itself does not exist without a signal.

#### The current definitions of uncertainty
Uncertainty is a concept that is famously hard to pin down, with constantly shifting definitions that mean it is quantified differently from author to author.

## EDA vs Communication
The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation.

There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication. Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals. Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation. Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community. Some authors simply confuse varianc for uncertainty. For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty. Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.  @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation. A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty. It is likely impossible to suppress all possible signals at once.

### No Such thing as an "uncertainty visualisation"
If uncertainty is a by-product of inference, a visualisation being an "uncertainty visualisation" does not exist.

What exactly is or is not an uncertainty visualisation is just as poorly defined as the definition of uncertainty, so authors seem to decide for themselves what is an uncertainty visualisation. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors.

@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

## Designing Visualisation for Signal supression

@@

# Visualising uncertainty as signal
If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables. This does not seem to be the case.
Uncertainty is noise, but most evaluation experiments measure it as signal.
The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.

@@

# How should we test uncertainty visualisations?
If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

>>>>

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
- Update background to reflect new motivation
-->

@@

There are many stages in our analysis that might benefit from the power of data visualisation, however this does not mean it is appropriately utilized or done with with effectiveness. Unlike an infographic, a statistical graphic is one of intent, and all enhancements made on a statistical graphic should be in pursuit of that goal (@Vanderplas2020). Visualization is an important step in exploratory data analysis (EDA) and communication. At the exploration stage, it is utilized to **learn** what is important about a data set while the communication stage is utilized to **tell** others about what we have learned. Each of these visualisation goals utilize an important attribute of visualisation. In both these cases, visualisations are powerful tools that can help us understand **how** our data or results might diverge from pre-conceived notions or what we expect.

@@

The review provided here attempts to address these issues by offering a novel perspective on the uncertainty visualisation problem. We start by discussing the purpose of uncertainty visualisations and why it has become it's own sub-field, distinct from "normal" information visualisation. Following this, we discuss what is an uncertainty definition and what differentiates it from a "normal" visualisation. Finally, we discuss how uncertainty visualisations can be evaluated and why evaluation of uncertainty visualisations can be more difficult than evaluations of "normal" visualisations.  By considering the *purpose* of visualizing uncertainty we avoid providing a simple list of the current methods, and instead consider if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology.

# Why is uncertainty special?
It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*. A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic. Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, median, maximum, or of any other statistic, however this field *does* implicitly treat them different. There is no such thing as "maximum visualiation" or "median visualisation" so there *must* be something about uncertainty that causes authors to believe it is special. Here we seek to answer this question.

### The purpose of uncertainty visualisations
Uncertainty visualisation papers usually provide one of two primary justifications for their existence. The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty. Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019]. The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate. Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].

@@

## What is signal-supression?
The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.

@@

@@

## Uncertainty visualisations vs information visualisation
If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables. This does not seem to be the case.
Uncertainty is noise, but most evaluation experiments measure it as signal.
The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.

@@

# What even is an uncertainty visualisation?
## The literatures definition of uncertainty visualisations
What exactly is an uncertainty visulisation is never explained and certainly does not have a consistent definition. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors.

### What is an uncertainty visualisation?
@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

### What is uncertainty?

Uncertainty is a concept that is famously hard to pin down, with constantly shifting definitions that mean it is quantified differently from author to author.

## What is an uncertainty visualisation?
The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication. Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.

 Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation. Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community. Some authors simply confuse varianc for uncertainty. For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty. Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.  @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation. A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty. It is likely impossible to suppress all possible signals at once.

# How do we evaluate uncertainty visualisations?
If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
